Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-19 treestats_paper.tex
edits until duality

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We use this correspondence to define a general framework that
efficiently computes single-site population genetic statistics
using the succinct tree sequence encoding of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two dual ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
under the infinite-sites model.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimized matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may encode interesting biological
signals.

@@

It was once a major undertaking to collect data sufficient to estimate a single summary
of the genetic relationships between the individuals of a
sample~\citep[e.g.][]{Kreitman1983-xr}.
% e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
% \citep{Kreitman1983-xr}.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes, both between
individuals~\citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}
and within
populations~\citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread}.
Computation is beginning to be a major problem:
projects such as UK Biobank~\citep{bycroft2018genome}
and gnomAD~\citep{karczewski2019variation} hold genetic data for
hundreds of thousands of samples at tens to hundreds of millions of variant sites.
Such large genotype matrices are extremely unwieldy and difficult to process.
Tools such as \texttt{hail} (\url{https://hail.is}) allow computations to be sharded
across many machines in parallel, thereby calculating statistics far more
quickly than is possible using single-computer methods such as
\texttt{plink}~\citep{purcell2007plink}. Computer time must still be
paid for, however, and running calculations concurrently on thousands of cores quickly
becomes expensive. Larger datasets still, consisting of millions of whole
genomes, are currently being collected, and it is clear that processing
and storing these genotype matrices will be very costly.

Fortunately, the genotype matrix is not the only way to encode genetic data.
Trees describe the ancestral relationships between
samples~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encode genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.

The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, while also
enabling efficient calculation of several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

The basic relationship between tree shape and summaries of genetic variation
is this: if mutations are neutral then the expected number that occur on a
given branch of a tree is proportional to the length of that branch.
This ``duality'' (we will use the term more precisely later) has been
used extensively in deriving properties of statistics in models of random mating
\citep[e.g.][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
as has the fact that it applies regardless of the underlying demographic model
\citep[e.g.][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
mcvean2002genealogical,ralph2019empirical}. We build on this central
insight to define a flexible and computationally efficient framework for
population genetic statistics on tree sequences.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarizing them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

\begin{figure}
    \centering
    \includegraphics{figures/divergence_diagram}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) lie in that subtree.
        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \label{fig:divergence_diagram}
    }
\end{figure}

The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to the framework described by \citet{fu1995statistical}.
To connect our  generalization to familiar ideas,
here is a reframing of \citeauthor{fu1995statistical}'s statistics.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

Our framework is more general as we can also define statistics that are
\emph{not} functions of the allele frequency spectrum.
Before we define the framework, we need some notation and definitions.
A tree sequence describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
Each haplotype, modern or ancestral, is associated with a \emph{node},
and the trees at each position along the genome have vertexes labeled by these nodes.
A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
that appear in the trees at every point along the genome.
The other (non-sample) nodes may not appear in every tree,
as for instance if a portion of an ancestor's genome was not inherited by any of the samples.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(\treeseq)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarizing we have access to more than one aspect of each subtree.
We will often be interested in statistics defined in terms of different
subsets of our samples (corresponding to, e.g., different populations); it
is useful to have some additional notation to define these statistics.
Given a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

Perhaps the most natural way to summarize weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us, for each node, what proportion of the sample's genomes were inherited from that node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the subtree weight of node $u$ in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree of $u$ in $T_k$.
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

\paragraph{Polarization}
The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
at a site, and we may define a \emph{polarized} node statistic without the second term from our previous definition:

@@

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position ---
i.e., the \emph{infinite sites model} ---
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry ---
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch in each tree
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}\label{eqn:branch_stat_tree}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.
An example of this correspondence between site and branch statistics
is shown in Figure~\ref{fig:branch_site_diagram}. Here, we
see how the $f_4$ site statistic assigns a weight to each mutation
depending on its frequency in each of the four sample sets,
and how the corresponding branch statistic assigns the same weight to those branches.

\begin{figure}
    \centering
    \includegraphics{figures/branch_site_diagram}
    \caption{
    Computing the $f_4(A,B;C,D)$ statistic on a tree.
    \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
    where $p_X$ is the proportion of the samples inheriting that mutation (colored stars) in set $X$
    (shown in labels at the bottom).
    Note that each of the mutations shown occurs at a distinct site, so there is no back mutation.
    \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
    to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
    is the sum over these weights, multiplied by the length of the branch
    (and the span of the tree, not depicted).
        \label{fig:branch_site_diagram}
    }
\end{figure}

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarized} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarized)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.

\begin{example}[Mean TMRCA] \label{ex:branch_diversity}
    If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
    then $f(u)$ gives the probability that the branch ancestral to $u$
    lies on the path from one of two randomly chosen samples from $S$
    on the path up to their most recent common ancestor (MRCA).
    Therefore, $\branch(f, \iw)$
    gives the mean total distance in the tree between two samples from $S$,
    averaged across the sequence.
    This is twice the mean TMRCA if the samples are all from the same time.
\end{example}

\begin{example}[Phenotypic correlation with pedigree] \label{ex:branch_correlation}
    If we take $\iw$ and $f$ as in the ``Phenotypic correlations'' example above,
    what does $\branch(f, \iw)$ tell us?
    The statistic tells us the \emph{expected} correlation between phenotype and any mutations
    appearing on the tree, which is a summary of how much local relatedness
    aligns with similarity in phenotype.
\end{example}

>>>>

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We use this correspondence to define a general framework that
efficiently computes single-site population genetic statistics
using the succinct tree sequence encoding of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two dual ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
under the infinite-sites model of mutation.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimized matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may encode interesting biological
signals.

@@

It was once a major undertaking to collect data sufficient to estimate a single summary
of the genetic relationships between the individuals of a
sample~\citep[e.g.,][]{Kreitman1983-xr}.
% e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
% \citep{Kreitman1983-xr}.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes, both between
individuals~\citep[e.g.,][]{browning2010highresolution,aguillon2017deconstructing}
and within
populations~\citep[e.g.,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread}.
Computation is beginning to be a major problem:
projects such as UK Biobank~\citep{bycroft2018genome}
and gnomAD~\citep{karczewski2019variation} hold genetic data for
hundreds of thousands of samples at tens to hundreds of millions of variant sites.
Such large genotype matrices are extremely unwieldy and difficult to process,
and computational complexity is often linear in the size of the matrix. 
Tools such as \texttt{hail} (\url{https://hail.is}) allow computations to be sharded
across many machines in parallel, thereby calculating statistics far more
quickly than is possible using single-computer methods such as
\texttt{plink}~\citep{purcell2007plink}. Computer time must still be
paid for, however, and running calculations concurrently on thousands of cores quickly
becomes expensive. Larger datasets still, consisting of millions of whole
genomes, are currently being collected, and it is clear that processing
and storing these genotype matrices will be very costly.

Fortunately, the genotype matrix is not the only way to encode genetic data:
an alternative would be to take advantage of the relatedness amongst
individuals and represent the data using trees.
Trees describe the ancestral relationships between
samples~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encode genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.

The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and has formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

The basic relationship between tree shape and summaries of genetic variation
is this: if mutations are neutral then the expected number that occur on a
given branch of a tree is proportional to the length of that branch.
This ``duality'' (we will use the term more precisely later) has been
used extensively in deriving properties of statistics in models of random mating
\citep[e.g.,][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
as has the fact that it applies regardless of the underlying demographic model
\citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
mcvean2002genealogical,ralph2019empirical}. We build on this central
insight to define a flexible and computationally efficient framework for
population genetic statistics on tree sequences.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
\autoref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarizing them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

\begin{figure}
    \begin{center}
    \includegraphics{figures/divergence_diagram}
    \end{center}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) lie in that subtree.
        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \label{fig:divergence_diagram}
    }
\end{figure}

The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to the framework described by \citet{fu1995statistical}.
To connect our  generalization to familiar ideas,
here is a reframing of \citeauthor{fu1995statistical}'s statistics.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

Our framework is more general as we can also define statistics that are
\emph{not} functions of the allele frequency spectrum, such as the correlation between genotype and phenotype.
Before we define the framework, we need some notation and definitions.
A tree sequence describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
Each haplotype, modern or ancestral, is associated with a \emph{node},
and the trees at each position along the genome have vertexes labeled by these nodes.
A tree sequence describes the relationships between a special set of nodes, the \emph{samples},
that appear in the trees at every point along the genome.
The other (non-sample) nodes may not appear in every tree,
as for instance if a portion of an ancestor's genome was not inherited by any of the samples.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(\treeseq)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome between $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarizing we have access to more than one aspect of each subtree.
We will often be interested in statistics defined in terms of different subsets of our samples;
it is useful to have some additional notation to define these statistics.
Given a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

Perhaps the most natural way to summarize weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us, for each node, what proportion of the sample's genomes were inherited from that node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the sum of the weights of all samples descending from node $u$
in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all remaining weights.
If this is computed over only a ``window'' of the genome
then $L_k$ is replaced by the length of the portion of that window that the tree $T_k$ extends for,
and $L$ is replaced by the length of the window.

\paragraph{Polarization}
The node statistic defined above is appropriate for statistics of all alleles at given variable site.
For other statistics, one is only concerned with the derived state(s) at a site,
and we define a \emph{polarized} node statistic without the second term from our previous definition:

@@

An alternative and possibly more familiar way to compute heterozygosity
would be to use a polarized Site statistic with summary function $f(x) = 2 x(n-x)/n(n-1)$.
However, this approach fails with more than two alleles per site.

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position ---
i.e., the \emph{infinite sites model} ---
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry ---
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a multiplicative factor of the mutation rate.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}\label{eqn:branch_stat_tree}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} \beta_{T}(u) \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $\beta_{T}(u)$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.
An example of this correspondence between site and branch statistics
is shown in Figure~\ref{fig:branch_site_diagram}. Here, we
see how the $f_4$ site statistic assigns a weight to each mutation
depending on its frequency in each of the four sample sets,
and how the corresponding branch statistic assigns the same weight to those branches.

\begin{figure}
    \centering
    \includegraphics{figures/branch_site_diagram}
    \caption{
    Computing the $f_4(A,B;C,D)$ statistic on a tree.
    \textbf{(a)} Each mutation is given weight equal to $(p_A - p_B)(p_C - p_D)$,
    where $p_X$ is the proportion of the samples inheriting that mutation (colored stars) in set $X$
    (shown in labels at the bottom).
    Note that each of the mutations shown occurs at a distinct site, so there is no back mutation.
    \textbf{(b)} Each branch is assigned weight equal to the weight that would be assigned
    to any mutation falling in it (colored lines), and then the branch $f_4$ statistic
    is the sum over these weights, multiplied by the length of the branch.
        \label{fig:branch_site_diagram}
    }
\end{figure}

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome, as we did for Site statistics.
We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[i,j)}
    &=
    \frac{1}{j-i} \sum_{k=1}^{n(\treeseq)} \ell_k(i,j) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(i,j)$ is the length of the region in $[i,j)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(i,j) = \max(0, \min(j,a_k) - \max(i,a_{k-1}))$).
The \emph{polarized} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarized)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[i,j)}$ is defined using $\branchp(f, \iw)_T$ as before.

\begin{example}[Mean TMRCA] \label{ex:branch_diversity}
    If we take $\iw$ and $f$ exactly as in the ``Nucleotide diversity'' example above,
    then $f(u)$ gives the probability that the branch ancestral to $u$
    lies on the path from one of two randomly chosen samples from $S$
    on the path up to their most recent common ancestor (MRCA).
    Therefore, $\branch(f, \iw)$
    gives the mean total distance in the tree between two samples from $S$,
    averaged across the sequence.
    This is twice the mean time back to the MRCA if the samples are all from the same time.
\end{example}

\begin{example}[Phenotypic correlation with pedigree] \label{ex:branch_correlation}
    If we take $\iw$ and $f$ as in the ``Phenotypic correlations'' example above,
    then $\branch(f, \iw)$ gives the \emph{expected} correlation between phenotype and any mutations
    appearing on the tree. This is a summary of how much local relatedness
    aligns with similarity in phenotype.
\end{example}
